VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms

### 2019-01-15 paper/main.tex
cleaning up some TODOs and fixing a couple names I've heard back from

\title[VarSight: Prioritizing Clinically Reported Variants]{VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms}
\author[Holt \textit{et~al}.]{James M. Holt\,$^{\text{\sfb 1,}*}$, Brandon Wilk\,$^{\text{\sfb 1}}$, Camille Birch\,$^{\text{\sfb 1}}$, Donna Brown\,$^{\text{\sfb 1}}$, Manavalan Gajapathy\,$^{\text{\sfb 1}}$, Alex Moss\,$^{\text{\sfb 1}}$, Nadiya Sosonkina\,$^{\text{\sfb 1}}$, Melissa Wilk\,$^{\text{\sfb 1}}$, Julie Anderson\,$^{\text{\sfb 1}}$, Jeremy Harris\,$^{\text{\sfb 1}}$, Jacob Kelly\,$^{\text{\sfb 1}}$, Fariba Shaterferdosian\,$^{\text{\sfb 1}}$, Angelina Uno-Antonison\,$^{\text{\sfb 1}}$, Arthur Weborg\,$^{\text{\sfb 1}}$, Undiagnosed Diseases Network and Elizabeth A. Worthey\,$^{\text{\sfb 1}}$}
\address{$^{\text{\sf 1}}$Software Development and Informatics, HudsonAlpha Institute for Biotechnology, Huntsville, 35806, USA % and \\
%$^{\text{\sf 2}}$Department, Institution, City, Post Code, Country.}
}

@@

Codicem also performs filtering for the analysts to reduce the number of variants that are viewed through a standard clinical analysis.  We used the latest version of their primary clinical filter to replicate the standard filtering process of the analysts.  In short, the filter requires the following for a variant to pass through the clinical filter: sufficient total read depth, sufficient alternate read depth, low population frequency, at least one predicted effect on a transcript, at least one gene-disease association, and to not be a known, common false-positive from sequencing.  For the specific details on the filter used, refer to TODO:XXX (I think the source code?).  In general, the filter reduces the number of variant from the order of millions to hundreds (anecdotally, roughly 200-400 variants per proband after filtering).

@@

For categorical data, we used a two step approach to cleaning the data: bin-count encoding and principal component analysis.  First, we chose to use a bin-count because there are many categories where multiple categorical labels may be present at different quantities.  For example, a single ClinVar variant may have multiple entries where different sites have selected different levels of pathogenicity.  In this situation, we desired to capture not only the categorical label as a feature, but also the number of times that label occurred in the annotations.  Second, we found that the bin-count encoding tended to create many extra features (one per category per annotation type) that were ignored and/or diluted the useful features from the pool of features.  To reduce this dilution, we used principal component analysis \citep{jolliffe2011} to reduce the dimensions of each category and stored at most two features per categorical feature.  For specifics on variance explained by the top two PCA components per category, please see Supplementary document XXX (TODO: are we going to make this or no?)

>>>>

\title[VarSight: Prioritizing Clinically Reported Variants]{VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms}
\author[Holt \textit{et~al}.]{James M. Holt\,$^{\text{\sfb 1,}*}$, Brandon Wilk\,$^{\text{\sfb 1}}$, Camille Birch\,$^{\text{\sfb 1}}$, Donna M. Brown\,$^{\text{\sfb 1}}$, Manavalan Gajapathy\,$^{\text{\sfb 1}}$, Alex Moss\,$^{\text{\sfb 1}}$, Nadiya Sosonkina\,$^{\text{\sfb 1}}$, Melissa A. Wilk\,$^{\text{\sfb 1}}$, Julie A. Anderson\,$^{\text{\sfb 1}}$, Jeremy Harris\,$^{\text{\sfb 1}}$, Jacob Kelly\,$^{\text{\sfb 1}}$, Fariba Shaterferdosian\,$^{\text{\sfb 1}}$, Angelina Uno-Antonison\,$^{\text{\sfb 1}}$, Arthur Weborg\,$^{\text{\sfb 1}}$, Undiagnosed Diseases Network and Elizabeth A. Worthey\,$^{\text{\sfb 1}}$}
\address{$^{\text{\sf 1}}$Software Development and Informatics, HudsonAlpha Institute for Biotechnology, Huntsville, 35806, USA % and \\
%$^{\text{\sf 2}}$Department, Institution, City, Post Code, Country.}
}

@@

Codicem also performs filtering for the analysts to reduce the number of variants that are viewed through a standard clinical analysis.  We used the latest version of their primary clinical filter to replicate the standard filtering process of the analysts.  In short, the filter requires the following for a variant to pass through the clinical filter: sufficient total read depth, sufficient alternate read depth, low population frequency, at least one predicted effect on a transcript, at least one gene-disease association, and to not be a known, common false-positive from sequencing.  In general, the filter reduces the number of variant from the order of millions to hundreds (anecdotally, roughly 200-400 variants per proband after filtering).  For the specific details on the filter used, please refer to the metadata available in the VarSight repository.

@@

For categorical data, we used a two step approach to cleaning the data: bin-count encoding and principal component analysis.  First, we chose to use a bin-count because there are many categories where multiple categorical labels may be present at different quantities.  For example, a single ClinVar variant may have multiple entries where different sites have selected different levels of pathogenicity.  In this situation, we desired to capture not only the categorical label as a feature, but also the number of times that label occurred in the annotations.  Second, we found that the bin-count encoding tended to create many extra features (one per category per annotation type) that were ignored and/or diluted the useful features from the pool of features.  To reduce this dilution, we used principal component analysis \citep{jolliffe2011} to reduce the dimensions of each category and stored at most two features per categorical feature.  %For specifics on variance explained by the top two PCA components per category, please see Supplementary document X (If we ever do this, it will be at the request of a user or reviewers, it doesn't seem super important at this juncture)
